Reconstructing Propositional Proofs in Type Theory

### 2017-10-23 paper/sections/proof-reconstruction/clausify.tex
[ paper ] Eq removed.
ruu

In the following \Metis \TSTP derivation, we see an example where
\clausify expands the formula \texttt{norm0} by using distributive
laws presented in conjunctive normal form conversion.

\begin{verbatim}
fof(a1, axiom, p => (q & r)).
...
fof(norm0, (~ p | (q & r)), inf(canonicalize, [a1])).
fof(norm1, ((~ p | q) & (~ p | r)), inf(clausify, [norm0])).
...
\end{verbatim}

@@

\begin{proof}
If $φ ≡ ψ$, $Γ ⊢ \fclausify~φ~ψ$ normalizes to $Γ ⊢ ψ$. The conclusion follows by applying the $\fsubst$ lemma.\\[2mm]
Otherwise, we use Theorem~\ref{thm:thm-reorder-and-or} and Theorem~\ref{thm:thm-cnf}.
\end{proof}

\begin{remark}
The \clausify rule is often preceded by the \canonicalize rule.
Both rules perform \emph{Clausification} that introduces
into the problem axioms or definitions in the domain of the solution.
The clausification algorithms is mainly described on paper by
\citeauthor{Sutcliffe1996}~\cite{Sutcliffe1996}.
\end{remark}

>>>>

In the following \Metis \TSTP derivation, we see an example where
\clausify expands the formula \texttt{norm0} by using distributive
laws included in the conjunctive normal form conversion.

\begin{verbatim}
fof(a1, axiom, p => (q & r)).
...
fof(norm0, ~ p | (q & r), inf(canonicalize, [a1])).
fof(norm1, (~ p | q) & (~ p | r), inf(clausify, [norm0])).
...
\end{verbatim}

@@

\begin{proof}
If $φ ≡ ψ$, $Γ ⊢ \fclausify~φ~ψ$ normalizes to $Γ ⊢ ψ$. The conclusion follows by applying the $\fsubst$ lemma. Otherwise, we use Theorem~\ref{lem:reorder-and-or} and Theorem~\ref{lem:cnf}.
\end{proof}

\begin{remark}
The \clausify rule is often preceded by \canonicalize.
Both rules perform \emph{Clausification} that introduces
into the problem axioms and definitions in the domain of the solution.
The clausification algorithms is mainly described on paper by
\citeauthor{Sutcliffe1996}~\cite{Sutcliffe1996}.
\end{remark}

### 2017-10-23 paper/sections/proof-reconstruction/resolve.tex
ruu

\begin{equation*}
 \label{eq:build-or}
  \begin{aligned}
  &\hspace{.495mm}\rm{build}_{∨} : \Prop → \Prop → \Prop\\
  &\begin{array}{llll}
    \rm{build}_{∨}&ψ~ψ &= ψ\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if }\rm{build}_{∨}~φ~ψ_{1} ≡ ψ_{1}\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if }\rm{build}_{∨}~φ~ψ_{2} ≡ ψ_{2}\\
    \rm{build}_{∨}&φ~ψ&=φ&\text{ otherwise.}
   \end{array}
  \end{aligned}
\end{equation*}

@@

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{sbuild}_{∨}~φ~ψ$, where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\rm{sbuild}_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{llll}
    \rm{sbuild}_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor(\rm{build}_{∨}~φ₁~ψ ∨ \rm{build}_{∨}~φ₂~ψ)&\\
    \rm{build}_{∨}&φ~ψ &= \rm{build}_{∨}~φ~ψ& \text{ otherwise.}
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

@@

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$, where, 
    \begin{align*}
      \begin{split}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                  &\text{and }ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

\begin{definition}[conjunct$_{\boldsymbol{∨}}$]
  \label{eq:conjunct-or-definition}
\begin{align*}
  \begin{split}
    &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
    &\fconjunct_{∨}~φ~ψ =
    \begin{cases}
      ψ, &\text{ if }φ ≡ ψ\\
      ψ, &\text{ if }\freorder_{∨}~φ~ψ ≡ ψ\\
      ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ \fconjunct_{∨}~φ~ψ₁ ≡ ψ₁,\\
        &\text{and } \freorder_{∨}~φ~ψ₂ ≡ ψ₂\\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ \fconjunct_{∨}~φ₁~ψ ≡ ψ\\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ \fconjunct_{∨}~φ₂~ψ ≡ ψ\\
      φ, &\text{ otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{definition}

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$.
\end{mainlemma}

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where
   \begin{align*}
      \begin{split}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ~=
        \begin{cases}
          ψ, &\text{ if } φ≡ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁~\\
             &\text{and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂~\\
          φ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ~ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

>>>>

\begin{equation*}
 \label{eq:build-or}
  \begin{aligned}
  &\hspace{.495mm}\rm{build}_{∨} : \Prop → \Prop → \Prop\\
  &\begin{array}{llll}
    \rm{build}_{∨}&ψ~ψ &= ψ\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if } ψ_{1} ≡ \rm{build}_{∨}~φ~ψ_{1}\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if } ψ_{2} ≡\rm{build}_{∨}~φ~ψ_{2}\\
    \rm{build}_{∨}&φ~ψ&=φ&\text{ otherwise.}
   \end{array}
  \end{aligned}
\end{equation*}

@@

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{sbuild}_{∨}~φ~ψ$, where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\rm{sbuild}_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{lll}
    \rm{sbuild}_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor(\rm{build}_{∨}~φ₁~ψ ∨ \rm{build}_{∨}~φ₂~ψ)\\
    \rm{build}_{∨}&φ~ψ &= \rm{build}_{∨}~φ~ψ
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

@@

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$, where, 
    \begin{align*}
      \begin{split}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                  &\text{and } ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$, where, 
  \begin{equation*}
\label{eq:conjunct-or-definition}
  \begin{split}
    &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
    &\fconjunct_{∨}~φ~ψ =
    \begin{cases}
      ψ, &\text{ if }φ ≡ ψ\\
      ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
      ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
        &\text{and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
      φ, &\text{ otherwise.}
    \end{cases}
  \end{split}
  \end{equation*}
\end{mainlemma}

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align*}
      \begin{split}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

### 2017-10-23 paper/sections/proof-reconstruction/strip.tex
ruu

\begin{mainlemma}
  \label{lem:inv-uh-lem}
Let $\text{n} : \Nat$ be the complexity measure of the \rm{\fuh₀} function
in~(\ref{eq:strip-unfixed}).
If $Γ ⊢ \fuh₁~φ~n$ then $Γ ⊢ φ$.
\end{mainlemma}

\begin{proof}
Use induction on the cases defined by the outcome of the
$\fuh$ function in~(\ref{eq:uh-structured}).
\begin{itemize}
  \item If $n = 0$, by definition we conclude $Γ ⊢ φ$.
  \item If $n = 1$, we apply the following theorem proved in~\cite{AgdaProp},
    \begin{equation*}
    \texttt{∧⇒\rm{-to-}⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
    \end{equation*}
  \item For $n > 1$, we use induction on the structure of the second
        argument.
\vskip 2mm
\begin{itemize}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fuh~(φ₁ ⇒ (φ₂ ⇒ φ₃))~(\suc~n)$}
  \RightLabel{by def.}
  \UnaryInfC{$Γ ⊢ \fuh~((φ₁ ∧ φ₂) ⇒ φ₃)~n$}
  \RightLabel{by hip.}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒\rm{-to-}⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)$.
\begin{equation*}
  \scalebox{0.95}{
  ($\mathcal{D}_1$)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh~(φ₁ ⇒ φ₂)~n ∧ \fuh~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₁}
    \UnaryInfC{$Γ ⊢ \fuh~(φ₁ ⇒ φ₂)~n$}
    \RightLabel{by hip.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$}
  \end{bprooftree}
  }
\end{equation*}

\begin{equation*}
  (\mathcal{D}_2)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh~(φ₁ ⇒ φ₂)~n ∧ \fuh~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₂}
    \UnaryInfC{$Γ ⊢ \fuh~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{by hip.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₃$}
    \end{bprooftree}
\end{equation*}

Finally, using the theorem \texttt{⇒∧⇒\rm{-to-}⇒∧} from~\cite{AgdaProp},
\begin{equation*}
  \texttt{⇒∧⇒\rm{-to-}⇒∧}\ :\ Γ ⊢ (φ₁ ⇒ φ₂) ∧ (φ₁ ⇒ φ₃) → Γ ⊢ φ₁ ⇒ (φ₂ ∧ φ₃),
\end{equation*}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}_1$}
  \AxiomC{$\mathcal{D}_2$}
  \RightLabel{∧-intro}
  \BinaryInfC{$Γ ⊢ (φ₁ ⇒ φ₂) ∧ (φ₁ ⇒ φ₃)$}
  \RightLabel{\tt ⇒∧⇒\rm{-to-}⇒∧}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ∧ φ₃)}
  \end{bprooftree}
\end{equation*}
\item Other cases are proved in a similar way.
\end{itemize}
\end{itemize}
\end{proof}

@@

The function defined above has termination problems since it is not
a structural recursion. Therefore, we have reformulated it
in~(\ref{eq:strip-fixed}) in order to solve such an issue following
the technique described in Section~\ref{ssec:structural-recursion}.

@@

$\bullet$~Case $φ ≡ ¬ ¬ φ₁$.
\begin{equation*}
  \begin{bprooftree}
 \AxiomC{$Γ ⊢ \fstrip~(¬¬φ₁)~(\suc~n)$}
  \RightLabel{by~(\ref{eq:uh-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip~φ₁~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ \fstrip~φ₁~n$}
  \RightLabel{by hip.}
  \UnaryInfC{$Γ ⊢ φ₁$}
  \end{bprooftree}
\end{equation*}

>>>>

\begin{mainlemma}
  \label{lem:inv-uh-lem}
Let $\text{n} : \Nat$ be the complexity measure of the $\fuh_0$ function
in~(\ref{eq:uh-structured}).
If $Γ ⊢ \fuh₁~φ~n$ then $Γ ⊢ φ$.
\end{mainlemma}

\begin{proof}
The proof is by induction on the cases defined by the outcome of the
$\fuh_{1}$ function in~(\ref{eq:uh-structured}).
\begin{itemize}
  \item If $n = 0$, by definition we conclude $Γ ⊢ φ$.
  \item If $n = 1$, we apply the following theorem proved in~\cite{AgdaProp},
    \begin{equation*}
    \texttt{∧⇒\rm{-to-}⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
    \end{equation*}
  \item For $n > 1$, we use induction on the structure of the second
        argument.
\vskip 2mm
\begin{itemize}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ⇒ φ₃))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:uh-structured})}
  \UnaryInfC{$Γ ⊢ \fuh_1~((φ₁ ∧ φ₂) ⇒ φ₃)~n$}
  \RightLabel{by hip.}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒\rm{-to-}⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)$.
\begin{equation*}
  \scalebox{0.95}{
  ($\mathcal{D}_1$)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ∧ φ₃))~(\suc~n)$}
    \RightLabel{by~(\ref{eq:uh-structured})}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n ∧ \fuh_1~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₁}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n$}
    \RightLabel{by hip.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$}
  \end{bprooftree}
  }
\end{equation*}

\begin{equation*}
  (\mathcal{D}_2)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ∧ φ₃))~(\suc~n)$}
    \RightLabel{by~(\ref{eq:uh-structured})}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n ∧ \fuh_1~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₂}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{by hip.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₃$}
    \end{bprooftree}
\end{equation*}

Now, using the theorem \texttt{⇒∧⇒\rm{-to-}⇒∧} from~\cite{AgdaProp},
\begin{equation*}
  \texttt{⇒∧⇒\rm{-to-}⇒∧}\ :\ Γ ⊢ (φ₁ ⇒ φ₂) ∧ (φ₁ ⇒ φ₃) → Γ ⊢ φ₁ ⇒ (φ₂ ∧ φ₃),
\end{equation*}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}_1$}
  \AxiomC{$\mathcal{D}_2$}
  \RightLabel{∧-intro}
  \BinaryInfC{$Γ ⊢ (φ₁ ⇒ φ₂) ∧ (φ₁ ⇒ φ₃)$}
  \RightLabel{\tt ⇒∧⇒\rm{-to-}⇒∧}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ∧ φ₃)}
  \end{bprooftree}
\end{equation*}
\item Other cases are proved in a similar way.
\end{itemize}
\end{itemize}
\end{proof}

@@

The function defined above is not a structural recursion function. Therefore, we have reformulated it in~(\ref{eq:strip-fixed}) in order to solve such an issue following the technique described in Section~\ref{ssec:structural-recursion}.

@@

$\bullet$~Case $φ ≡ ¬ ¬ φ₁$.
\begin{equation*}
  \begin{bprooftree}
 \AxiomC{$Γ ⊢ \fstrip~(¬¬φ₁)~(\suc~n)$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip~φ₁~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ \fstrip~φ₁~n$}
  \RightLabel{by hip.}
  \UnaryInfC{$Γ ⊢ φ₁$}
  \end{bprooftree}
\end{equation*}
